\chapter{Relations and Functions}\label{ch:functions}
\section{Relations}

\begin{definition}
  A (binary) relation \(R\) between two sets \(A, B\) is the 3-tuple of those
  two sets and any subset of their Cartesian product:
  \begin{align*}
    R \defeq (A, B, R') & \text{where} & R' \subseteq A \times B
  \end{align*}
  Which means that a relation between two sets is specified by those two sets
  and a set of pairs, in which the first element of each pair is from one set
  and the second element of each pair is from the other.

  The set \(A\) is then called the \emph{domain} of \(R\), and \(B\) is called
  its \emph{codomain}.
\end{definition}

A relation \emph{relates} objects from one set to objects from another (or possibly the same) set.

To express that a given pair is in such a relation, we may use
any of the following notations:
\begin{center}
\begin{tabular}{c c}
  \((a, b) \in R\) & \text{explicit set notation}\\
  \(R(a,b)\) & \text{prefix notation}\\
  \(a R b\) & \text{infix notation}
\end{tabular}
\end{center}

\begin{example}
  [Example of a relation]
  Let \(A=\NN, B=\NN\).
  Then \(<\) defines a relation from \(A\) to \(B\).\\
  For instance, \((1, 2) \in <\), or as this is more commonly denoted:
  \(1 < 2\). \\
  However, \((1, 0) \notin <\), meaning \( 1 \not< 0\).
\end{example}

\subsection{Uniqueness Properties}

A relation is called \emph{left-unique} if every element in its domain
is related to \emph{at most one} element of its codomain.
Similarly, a relation is called \emph{right-unique} if every element in the
codomain is related to \emph{at most one} element in the domain.

\subsection{Totality Properties}

A relation is called \emph{left-total} if every element in its domain
is related to \emph{at least one} element of its codomain.
Again, the term \emph{right-total} is used to describe the same idea
the other way around: every element of the codomain of a right-total relation
is related to \emph{at least one} element of its domain.

\section{Functions}

\begin{definition}
  A \emph{function} is a relation that is right-unique and left-total.
  We usually denote this as
  \begin{align*}
    f \colon A \to B
  \end{align*}
  meaning \emph{``Let \(f\) be a function from \(A\) to \(B\)''}.
  This means that \(f\) assigns each \(a \in A\) \emph{exactly one}
  \(b \in B\), though it may repeatedly assign the same elements of \(B\),
  or not assign some elements from \(B\) to any elements in \(A\) at all.
\end{definition}

\begin{definition}
  A function is called \emph{injective} if it is also left-unique, meaning it
  only assigns each element of the codomain \textit{at most once}.\\
  A function is called \emph{surjective} if it is also right-total, meaning it
  assigns each element of the codomain \textit{at least once}.\\
  A function is called \emph{bijective}, or \textit{one-to-one},
  if it is both injective and surjective.
\end{definition}

\begin{tikzpicture}

% Y coordinates for 3 points each side
\def\domainY{1.5,0,-1.5}
\def\codomainY{1.5,0,-1.5}

\ovalrelation{0cm}{0cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/2}
  {A \textit{left-unique}, \textit{right-unique} relation}

\ovalrelation{7cm}{0cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,1/2}
  {A \textit{left-unique} relation}

\ovalrelation{0cm}{-6cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/2,2/3}
  {A \textit{left-unique}, \textit{right-total} relation}

\ovalrelation{7cm}{-6cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/2,3/2}
  {\shortstack{A \textit{function} \\ (which is not injective or surjective)}}

\ovalrelation{0cm}{-12cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/2,3/1}
  {An \textit{injective} function}

\ovalrelation{7cm}{-12cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/1,3/2}
  {A \textit{surjective} function}

\ovalrelation{3.5cm}{-18cm}
  {1.5,0,-1.5}
  {1.5,0,-1.5}
  {1/1,2/2,3/3}
  {A \textit{bijective} function}

\end{tikzpicture}
